---
title: "Defining hill farming"
author: "Mike Spencer"
date: "23 May 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F)
library(rgdal)
library(broom)
library(tidyverse)
library(viridis)
library(knitr)
```


## Introduction

### Purpose

This document is part of a project evaluating the importance of hill farming in Scottish communities. There is no existing classification of hill farming in Scotland, therefore an index of hill farming needed to be created to evalute its contribution. We chose to do this at an agricultural parish level using datasets describing landscape.

### Output level

Results will be aggregated to agricultural parishes, as this administrative level is widely understood and is appropriate for anonymising data.
The 2016 agricultural parish data were downloaded from: <https://data.gov.uk/dataset/939fdd5e-7322-4ab7-9dc9-bbfc538c4477/agricultural-parishes>.

```{r agricultural_parishes, fig.cap="2016 agricultural parishes"}
parishes = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "simplified_parishes")

Scotland = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "Scotland") %>% 
   tidy()

parishes$id = row.names(parishes)
parishes = tidy(parishes) %>% 
   left_join(parishes@data)

ggplot(parishes, aes(long, lat, group=group)) +
   geom_polygon(colour="grey30", fill="grey90", size=0.05) +
   coord_equal() +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```

The above map shows the `r length(unique(parishes$PARCode))` agricultural parishes in Scotland.
These vary in size between `r round(min(parishes$Shape_Area)/1e+06, 1)` and `r round(max(parishes$Shape_Area)/1e+06)` km².

## Defining hill farming

A simple view of hill farming is of an activity which takes place in hills, but how are hills defined?
Measuring hilliness by elevation would diminish the challenges of farming smaller hill areas like the Pentlands and Campsies, because they are a third of the height of the biggest mountains in Scotland.
Incorporating slope into a measure of hill farming would help overcome using only elevation as a measure of hilliness because slopes can be steep without hills needing to be high.

There are many datasets available which we can use to further form an idea of hill farming areas. We have collected data from a variety of sources to help define hill farming:

* Elevation (including derived slope): <https://www.ordnancesurvey.co.uk/business-and-government/products/terrain-50.html>
* Scottish Natural Heritage (SNH) Landscape Character Assessment (LCA): <https://gateway.snh.gov.uk/natural-spaces/>
* SNH Carbon and Peatland map: <https://gateway.snh.gov.uk/natural-spaces/>
* James Hutton Institute (JHI) Land capability for agriculture: <http://www.hutton.ac.uk/learning/natural-resource-datasets>
* Agricultural census common and rough grazing: RESAS

```{r data load}
restrictions = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/parish_restrictions.csv") %>% 
   mutate_all(funs(replace(., is.na(.), 0))) %>% 
   mutate(peat_prop=as.numeric(peat_prop))

parishes = parishes %>% 
   left_join(restrictions, by=c("PARCode"="PARCode", "PARName"="PARName")) %>% 
   mutate_all(funs(replace(., is.na(.), 0)))
```

### Definition method

Data described in the previous section were available in a range of formats, including raster (grid), vector (polygons) and plain text.
In order to assess all these data at an agricultural parish level, these data were aggregated and converted to a common format.
Following this process data were converted to a proportion of hilliness in each parish, e.g. for rural urban classification the proportion of each parish classified as rural was calculated.

Exact specification for this process is detailed in:

* `grass_read.R` which reads files into GRASS GIS
* `grass_analysis.R` clips polygon data to agricultural parishes and measures subsequent polygon sizes. Terrain values (elevation and slope) are also extracted from raster data.
* `read_hill_class.R` combines data from the GRASS analysis and aggregates to parish levels. Crucially, at a parish level proportion of land is reported as being less viable, i.e. higher proportions mean a parish is more likely to be hill farming.

```{r functions}
plot.map = function(i, tit){
   ggplot(parishes, aes(long, lat, group=group)) +
      geom_polygon(aes_string(fill=i)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

panel.cor = function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr = par("usr"); on.exit(par(usr))
    par(usr=c(0, 1, 0, 1))
    r = cor(x, y)
    txt = format(c(r, 0.123456789), digits = digits)[1]
    txt = paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor = 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```


### Terrain

The Ordnance Survey Terrain 50 digital elevation model was used to get terrain data for each parish.
This gridded model has a 50 m cell resolution and each cell is attributed an average height of land contained within it.
This elevation model was used to calculate slope, also on a 50 m grid.
For the slope and elevation datasets summary values were extracted for each agricultural parish; these values were minimum, mean, median and maximum.

#### Elevation

The four measures of elevation were compared in order to select a single indicator.
These comparisons are shown below, with the numeric values being Pearson correlation coefficients.

```{r elevation_pairs, fig.height=8, fig.width=8, fig.cap="Comparisons of elevation measures"}
restrictions %>% 
   select(elev__minimum, elev__average, elev__median, elev__maximum) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Elevation measure comparisions")
```

Mean (average), median and maximum elevation are closely correlated, but there is variability between each of these and minimum elevation.
This is due to within-parish elevation variation, e.g. some parishes will have a range from sea level to mountain top, but other parishes a occupy more level terrain.
The elevation range within a parish could have been used, but this would largely duplicate slope, i.e. a big elevation range within a parish should produce a steeper slope.

Mean parish elevation has been chosen as a single measure of parish elevation.
This because it is a central measure which captures the variation of the minimum, median and maximum best (correlations of 0.77, 0.99 and 0.94 respectively).

```{r elevation, fig.cap="Map of mean parish elevation"}
plot.map("elev__average", "Elevation\nmean (m)")
```


#### Slope

As for elevation, slope measures were compared to each other (below).
This plot matrix shows scatter plots between the measures in the lower left and Pearson's correlations in the upper right.

```{r slope_pairs, fig.height=8, fig.width=8, fig.cap="Comparison of slope measures"}
restrictions %>% 
   select(slope__minimum, slope__average, slope__median, slope__maximum) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Slope measure comparisions")
```

As with elevation, the average (mean and median) slope measures correlate well.
There is greater variability when compared to maximum slopes, but the largest variation is when comparing to minimum slopes.

A mean (average) slope has been used as a single parish slope indicator because it has the best correlation with median or maximum slope (0.98 and 0.71 respectively).

```{r slope, fig.cap="Map of mean parish slope"}
plot.map("slope__average", "Slope\nmean\n(degrees)")
```

### Grazing

#### Common

The common grazing data were supplied by RESAS at a parish level in a csv format, so needed no format conversion or aggregation preparation.
Common grazing data were sent with units of Hectares per parish, these were converted to m² before deriving proportion of common grazing within each parish.
Common grazing is most dominant in the Outter Hebridies and North West Scotland.

```{r common_grazing, fig.cap="Map of parish common grazing proportion"}
plot.map("c_grazing_prop", "Common\ngrazing\nproportion")
```

#### Rough

Rough grazing areas were taken from the June Agricultural Census (2011).

```{r rough_grazing, fig.cap="Map of parish common grazing proportion"}
plot.map("r_grazing_prop", "Rough\ngrazing\nproportion")
```



### SNH LCA Landscape Character Assessment

The SNH landscape character assessment has three levels.
Level 1 has the most different landscape classes and level 3 has the least.
As this work is concerned with aggregating data to define parishes, level 3 data were used as a hill farming indicator.
Level 3 data were split as follows to define hill farming areas:

```{r SNH_lca_character}
x = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/LCA_SCOTLAND_parishes.csv") %>% 
   select(a_LEVEL_3) %>% 
   distinct()

y = c("Flat or Rolling, Smooth or Sweeping, Extensive, High Moorlands of the Highlands and Islands",
                           "Inland Loch",
                           "Highland Straths",
                           "Moorland Transitional Landscapes of the Highlands and Islands",
                           "High, Massive, Rolling, Rounded Mountains of the Highlands and Islands",
                           "High Massive Mountain Plateau of the Cairngorms",
                           "Smooth Upland Moorland Hills",
                           "Highland Foothills",
                           "Upland Igneous and Volcanic Hills The Ochil, Sidlaw, Cleish and Lomond Hills",
                           "High, Massive, Rugged, Steep-Sided Mountains of the Highlands and Islands",
                           "Sea Lochs of the Highlands and Islands",
                           "Highland and Island Rocky Coastal Landscapes",
                           "Peatland Landscapes  of the Highlands and Islands",
                           "Rocky Moorlands of the Highlands and Islands",
                           "Rugged, Craggy Upland Hills and Moorlands of the Highlands, including the Trossachs",
                           "Low Coastal Hills of the Highlands and Islands",
                           "Coastal Hills Headlands Plateaux and Moorlands",
                           "Lowland Hills",
                           "Foothills and Pronounced Hills",
                           "Upland Hills, The Southern Uplands and Cheviots",
                           "High Plateau Moorlands",
                           "Rugged Granite Uplands",
                           "Rocky Volcanic Islands",
                           "Rugged Moorland Hills",
                           "Upland Fringe Moorland",
                           "Upland Hills, The Lammemuir, Pentland and Moorfoot Hills",
                           "Rocky Coasts Cliffs and Braes of the Lowlands",
                           "Knock or Rock and Lochan of the Islands",
                           "Highland Cnocan")

data.frame(y) %>% 
   kable(col.names="Hill farming classes")

data.frame(x$a_LEVEL_3[!x$a_LEVEL_3 %in% y]) %>%
   kable(col.names="Not hill farming classes")
```

As can be seen, some of these descriptions are a little ambiguous.
For example "Upland Basin" describes both a hill and glen feature.

Following this separation of landscape classes, the proportions in each parish were calculated, these are mapped below.
The upland areas of the Cairngorms, Western Highlands and North West Scotland dominate this map.

```{r SNH_lca, fig.cap="Map of parish SNH upland landscape proportion"}
plot.map("landscape_class_prop", "SNH\nUpland\nLandscape\nProportion")
```


### Carbon and peatland

SNH produce a map of carbon rich soils and peatland, these are polygon data with each polygon taking a categorical variable.
These classifications were split into those which might indicate hill farming and those which do not.
These classifications are listed below:

```{r SNH_peat_areas}
x = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/PEAT_SCOTLAND_parishes.csv") %>% 
   select(a_PRIMARY_LA) %>% 
   distinct()

y=c("Peaty gleys", "Peaty gleyed podzols", "Scree", "Subalpine podzols", "Dystrophic blanket peat", "Alpine podzols", "Dystrophic basin peat", "Peaty rankers", "Undifferentiated basin peats", "Dystrophic semi-confined peat", "Rankers",  "Podzolic rankers", "Peaty alluvial soils", "Shingle", "Undifferentiated semi-confined peats", "Subalpine (Orohemiarctic) podzols", "Rock", "Peaty podzols", "Alpine (Oroarctic) podzols", "Alpine (Oroarctic) gleys", "Brown rankers")

data.frame(peat_areas=y) %>% 
   kable(col.names="Hill farming classes")

data.frame(not_peat_areas=x$a_PRIMARY_LA[!x$a_PRIMARY_LA %in% y]) %>%
   kable(col.names="Not hill farming classes")
```

Those believed to indicate hill farming constraints were used to calculate the proportion of agricultural parish within them and are plotted below.
North West Scotland appears as the highest proportion using these measures.

```{r SNH_peat, fig.cap="Map of parish SNH peat map proportion"}
plot.map("peat_prop", "Peat\nProportion")
```


### JHI LCA Land capability for agriculture

The James Hutton Institute produce a land capability index for agriculture based on soil, climate and relief.
We downloaded the 1:250k resolution version, as it covers the whole of Scotland.
The vector dataset classifies areas of Scotland by their agricultural capability, these classes are:

* Class 1. Land capable of producing a very wide range of crops with high yields
* Class 2. Land capable of producing a wide range of crops with yields less high than Class 1.
* Class 3. Land capable of producing good yields from a moderate range of crops.
* Class 4. Land capable of producing a narrow range of crops.
* Class 5. Land suited only to improved grassland and rough grazing.
* Class 6. Land capable only of use as rough grazing.
* Class 7. Land of very limited agricultural value.
* Classes >10. Urban areas, etc..

Polygons classed as 6 or 7 were aggregated and the proportion of agricultural parishes lying within them calculated.
These are shown in the following map, which has the highest proportions in the central and western Highlands.

```{r JHI_lca, fig.cap="Map of parish JHI agricultural capability proportion"}
plot.map("land_cap_ag_prop", "Agriculture\n6 & 7 class\nProportion")
```



## Score and rank for the hilliness of each parish

The previous section describes a range of indices which can indicate hill farming areas.
We have combined to produce a single index of hill farming areas.

The majority of the indicies have maximum values of one, because they show proportions.
However, elevation and slope have maximum values above one.
To create an unweighted index slope and elevation have been normalised by dividing values by their series maximums.
These individual indicies were then summed for each parish to create a single index of hill farming for Scotland.

The results of this summation can be seen in the map below.
The Cairngorms score highest, with areas in the central and western Highlands, North West Scotland and the Outer Hebridies also scoring highly.

```{r score, fig.cap="Map of parish hill farming index"}
score = restrictions %>% 
   select(-PARName,
          -elev__minimum, -elev__median, -elev__maximum,
          -slope__minimum, -slope__median, -slope__maximum) %>% 
   mutate(elev__average=elev__average/max(elev__average),
          slope__average=slope__average/max(slope__average)) %>% 
   gather(variable, value, -PARCode) %>% 
   group_by(PARCode) %>% 
   summarise(score=sum(value))

parishes = parishes %>% 
   left_join(score)

parishes %>% 
   ggplot(aes(long, lat, fill=score, group=group)) +
   geom_polygon() +
   coord_equal() +
   scale_fill_viridis(option="magma") +
   labs(fill="Hill\nscore") +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```

The index of hill farming was ranked to show relative positions of each parish.
These results are below:

```{r rank, fig.cap="Map of parish hill farming index rank"}
x = score %>% 
   mutate(rank=rank(desc(score))) %>% 
   select(-score)

parishes = parishes %>% 
   left_join(x)

ggplot(parishes, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=rank)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=-1) +
      labs(fill="Hilliness\nrank") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
```

The top 15 ranked parishes are:

```{r rank 20}
score %>% 
   left_join(select(restrictions, PARCode, PARName)) %>% 
   top_n(15, wt=score) %>% 
   select(PARName)
```


## Group parishes into similar levels

We can use the individual indicies to group similar parishes together.
First, we can inspect the relationships between indicies in the plot matrix below.
Numbers in the upper right panels are Pearson correlations.

```{r proportion_pairs, fig.height=10, fig.width=10, fig.cap="Comparison of hill farming indices by parish"}
restrictions %>% 
   select(elev__average, slope__average, r_grazing_prop, c_grazing_prop, landscape_class_prop, land_cap_ag_prop, peat_prop) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Hilliness indicator comparisons")
```

The most well correlated variables are the JHI land capability indicies and the SNH Peat and Carbon map.
These indicies use JHI soil data, so relationships are expected.

We can use a k-means model to group similar parishes together.
K-means works by looking for common averages to cluster data points around.
The following sections detail this work.

### Build k-means model

A k-means model was built using standarised elev__average, slope__average, grazing_prop, landscape_class_prop, land_cap_ag_prop, land_cap_forest_prop, wildland_prop, rural_prop, NNR_prop and peat_prop variables.
The optimum number of cluster were determined using the elbow method, where we try to minimise the total intra-cluster variation.
In this case, the below plot indicates two clusters may be appropriate, however these data are a little messy.

```{r kmean build, fig.cap="Elbow plot for determining optimal cluster number"}
x = restrictions %>% 
   select(elev__average, slope__average, c_grazing_prop, r_grazing_prop, landscape_class_prop, land_cap_ag_prop, peat_prop)

# Standardise values
x = scale(x)

# Determine number of clusters
wss = (nrow(x) -1) * sum(apply(x, 2, var))
for (i in 2:10){
   wss[i] = sum(kmeans(x, centers=i)$withinss)
}

plot(1:10, wss, type="b", main="Required number of clusters", xlab="Number of Clusters", ylab="Within groups sum of squares")
```

```{r kmean run}
set.seed(224)

fit2 = kmeans(x, 2) # 2 cluster solution

# append cluster assignment
restrictions = data.frame(restrictions, cluster2=fit2$cluster)
```

These two clusters can then be mapped to our parishes as two levels: hill farming areas and not hill farming areas (below).
As we expect, these are largely similar to the hill farming index, but partitioned into two groups.

```{r kmean_map, fig.cap="Map of parishes grouped by hill farming indices"}
parishes = parishes %>% 
   left_join(select(restrictions, PARCode, cluster2))

parishes %>% 
   ggplot(aes(long, lat, group=group)) +
   geom_polygon(aes(fill=as.character(cluster2))) +
   geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
   coord_equal() +
   scale_fill_brewer(palette="Dark2") +
   labs(fill="Hill or not") +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```

We can see the difference in hill farming index against the two cluster groups below:

```{r cluster index, fig.cap="Plot comparing hill farming index to clusters of hill farming"}
parishes %>% 
   group_by(PARCode) %>% 
   summarise(score=unique(score),
             cluster=unique(cluster2)) %>% 
   ggplot(aes(as.factor(cluster), score, fill=as.factor(cluster))) +
   geom_boxplot() +
   scale_fill_brewer(palette="Dark2") +
   labs(x="Cluster",
        y="Hill farming index",
        fill="Hilliness\ncluster") +
   theme_bw() +
   theme(line=element_blank())
```
