---
title: "Defining hill farming"
author: "Mike Spencer"
date: "16 April 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F)
library(rgdal)
library(broom)
library(tidyverse)
library(viridis)
library(knitr)
```


## Introduction

### Purpose

This document is part of a project evaluating the importance of hill farming in Scottish communities. There is no existing classification of hill farming in Scotland, therefore an index of hill farming needed to be created to evalute its contribution. We chose to do this at an agricultural parish level using datasets describing landscape.

### Output level

Results will be aggregated to agricultural parishes, as this administrative level is widely understood and is appropriate for anonymising data.
The 2016 agricultural parish data were downloaded from: <https://data.gov.uk/dataset/939fdd5e-7322-4ab7-9dc9-bbfc538c4477/agricultural-parishes>.

```{r agricultural_parishes, fig.cap="2016 agricultural parishes"}
parishes = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "simplified_parishes")

Scotland = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "Scotland") %>% 
   tidy()

parishes$id = row.names(parishes)
parishes = tidy(parishes) %>% 
   left_join(parishes@data)

ggplot(parishes, aes(long, lat, group=group)) +
   geom_polygon(colour="grey30", fill="grey90", size=0.05) +
   coord_equal() +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```

The above map shows the `r length(unique(parishes$PARCode))` agricultural parishes in Scotland.
These vary in size between `r min(parishes$Shape_Area)/1e+06` and `r max(parishes$Shape_Area)/1e+06` km².

## Defining hill farming

A simple view of hill farming is of an activity which takes place in hills, but how are hills defined?
Measuring hilliness by elevation would diminish the challenges of farming smaller hill areas like the Pentlands and Campsies, because they are a third of the height of the biggest mountains in Scotland.
Incorporating slope into a measure of hill farming would help overcome using only elevation as a measure of hilliness because slopes can be steep without hills needing to be high.

There are many datasets available which we can use to further form an idea of hill farming areas. We have collected data from a variety of sources to help define hill farming:

* Urban/rural classification: <http://www.gov.scot/Publications/2018/03/6040/downloads>
* Elevation (including derived slope): <https://www.ordnancesurvey.co.uk/business-and-government/products/terrain-50.html>
* Scottish Natural Heritage (SNH) Landscape Character Assessment (LCA): <https://gateway.snh.gov.uk/natural-spaces/>
* SNH Wild land: <https://gateway.snh.gov.uk/natural-spaces/>
* SNH National Nature Reserves (NNR): <https://gateway.snh.gov.uk/natural-spaces/>
* SNH Carbon and Peatland map: <https://gateway.snh.gov.uk/natural-spaces/>
* James Hutton Institute (JHI) Land capability for agriculture: <http://www.hutton.ac.uk/learning/natural-resource-datasets>
* JHI Land capability for forestry: <http://www.hutton.ac.uk/learning/natural-resource-datasets>
* Agricultural census common grazing: RESAS

```{r data load}
restrictions = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/parish_restrictions.csv") %>% 
   mutate_all(funs(replace(., is.na(.), 0)))

parishes = parishes %>% 
   left_join(restrictions, by=c("PARCode"="PARCode", "PARName"="PARName")) %>% 
   mutate_all(funs(replace(., is.na(.), 0)))
```

### Definition method

Data described in the previous section were available in a range of formats, including raster (grid), vector (polygons) and plain text.
In order to assess all these data at an agricultural parish level, these data were aggregated and converted to a common format.
Following this process data were converted to a proportion of hilliness in each parish, e.g. for rural urban classification the proportion of each parish classified as rural was calculated.

Exact specification for this process is detailed in:

* `grass_read.R` which reads files into GRASS GIS
* `grass_analysis.R` clips polygon data to agricultural parishes and measures subsequent polygon sizes. Terrain values (elevation and slope) are also extracted from raster data.
* `read_hill_class.R` combines data from the GRASS analysis and aggregates to parish levels. Crucially, at a parish level proportion of land is reported as being less viable, i.e. higher proportions mean a parish is more likely to be hill farming.

```{r functions}
plot.map = function(i, tit){
   ggplot(parishes, aes(long, lat, group=group)) +
      geom_polygon(aes_string(fill=i)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

panel.cor = function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr = par("usr"); on.exit(par(usr))
    par(usr=c(0, 1, 0, 1))
    r = cor(x, y)
    txt = format(c(r, 0.123456789), digits = digits)[1]
    txt = paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor = 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```


### Terrain

The Ordnance Survey Terrain 50 digital elevation model was used to get terrain data for each parish.
This gridded model has a 50 m cell resolution and each cell is attributed an average height of land contained within it.
This elevation model was used to calculate slope, also on a 50 m grid.
For the slope and elevation datasets summary values were extracted for each agricultural parish; these values were minimum, mean, median and maximum.

#### Elevation

The four measures of elevation were compared in order to select a single indicator.
These comparisons are shown below, with the numeric values being Pearson correlation coefficients.

```{r elevation_pairs, fig.height=8, fig.width=8}
restrictions %>% 
   select(elev__minimum, elev__average, elev__median, elev__maximum) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Elevation measure comparisions")
```

Mean (average), median and maximum elevation are closely correlated, but there is variability between each of these and minimum elevation.
This is due to within-parish elevation variation, e.g. some parishes will have a range from sea level to mountain top, but other parishes a occupy more level terrain.
The elevation range within a parish could have been used, but this would largely duplicate slope, i.e. a big elevation range within a parish should produce a steeper slope.

Mean parish elevation has been chosen as a single measure of parish elevation.
This because it is a central measure which captures the variation of the minimum, median and maximum best (correlations of 0.77, 0.99 and 0.94 respectively).

```{r elevation}
plot.map("elev__average", "Elevation\nmean (m)")
```


#### Slope

As for elevation, slope measures were compared to each other (below).
This plot matrix shows scatter plots between the measures in the lower left and Pearson's correlations in the upper right.

```{r slope_pairs, fig.height=8, fig.width=8}
restrictions %>% 
   select(slope__minimum, slope__average, slope__median, slope__maximum) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Slope measure comparisions")
```

As with elevation, the average (mean and median) slope measures correlate well.
There is greater variability when compared to maximum slopes, but the largest variation is when comparing to minimum slopes.

A mean (average) slope has been used as a single parish slope indicator because it has the best correlation with median or maximum slope (0.98 and 0.71 respectively).

```{r slope}
plot.map("slope__average", "Slope\nmean\n(degrees)")
```


### Urban rural

Agricultural parishes cover the entire land mass of Scotland, so some are predominantly urban areas.
Below is a map showing the proportion of each parish which is classified as rural in the Scottish Government two fold urban-rural classification.

```{r urban-rural_map}
plot.map("rural_prop", "Rural\nProportion")
```

Urban-rural classification does not make a distiction between upland and lowland areas, but a remoteness index of how long it takes to travel (by driving) to the nearest town.
Hence, it does not represent the ease or challenge of farming in a given parish.
It does, however give an indication of proximity to market.


### Common grazing

The common grazing data were supplied by RESAS at a parish level in a csv format, so needed no format conversion or aggregation preparation.
Common grazing data were sent with units of Hectares per parish, these were converted to m² before deriving proportion of common grazing within each parish.
Common grazing is most dominant in the Outter Hebridies and North West Scotland.

```{r common_grazing}
plot.map("grazing_prop", "Common\ngrazing\nproportion")
```


### SNH LCA Landscape Character Assessment

The SNH landscape character assessment has three levels.
Level 1 has the most different landscape classes and level 3 has the least.
As this work is concerned with aggregating data to define parishes, level 3 data were used as a hill farming indicator.
Level 3 data were split as follows to define hill farming areas:

```{r SNH_lca_character}
x = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/LCA_SCOTLAND_parishes.csv") %>% 
   select(a_LEVEL_3) %>% 
   distinct()

y = c("Flat or Rolling, Smooth or Sweeping, Extensive, High Moorlands of the Highlands and Islands",
                           "Inland Loch",
                           "Highland Straths",
                           "Moorland Transitional Landscapes of the Highlands and Islands",
                           "High, Massive, Rolling, Rounded Mountains of the Highlands and Islands",
                           "High Massive Mountain Plateau of the Cairngorms",
                           "Smooth Upland Moorland Hills",
                           "Highland Foothills",
                           "Upland Igneous and Volcanic Hills The Ochil, Sidlaw, Cleish and Lomond Hills",
                           "High, Massive, Rugged, Steep-Sided Mountains of the Highlands and Islands",
                           "Sea Lochs of the Highlands and Islands",
                           "Highland and Island Rocky Coastal Landscapes",
                           "Peatland Landscapes  of the Highlands and Islands",
                           "Rocky Moorlands of the Highlands and Islands",
                           "Rugged, Craggy Upland Hills and Moorlands of the Highlands, including the Trossachs",
                           "Low Coastal Hills of the Highlands and Islands",
                           "Coastal Hills Headlands Plateaux and Moorlands",
                           "Lowland Hills",
                           "Foothills and Pronounced Hills",
                           "Upland Hills, The Southern Uplands and Cheviots",
                           "High Plateau Moorlands",
                           "Rugged Granite Uplands",
                           "Rocky Volcanic Islands",
                           "Rugged Moorland Hills",
                           "Upland Fringe Moorland",
                           "Upland Hills, The Lammemuir, Pentland and Moorfoot Hills",
                           "Rocky Coasts Cliffs and Braes of the Lowlands",
                           "Knock or Rock and Lochan of the Islands",
                           "Highland Cnocan")

data.frame(y) %>% 
   kable(col.names="Hill farming classes")

data.frame(x$a_LEVEL_3[!x$a_LEVEL_3 %in% y]) %>%
   kable(col.names="Not hill farming classes")
```

As can be seen, some of these descriptions are a little ambiguous.
For example "Upland Basin" describes both a hill and glen feature.

Following this separation of landscape classes, the proportions in each parish were calculated, these are mapped below.
The upland areas of the Cairngorms, Western Highlands and North West Scotland dominate this map.

```{r SNH_lca}
plot.map("landscape_class_prop", "SNH\nUpland\nLandscape\nProportion")
```

### SNH Wild land

SNH's Wild land data are binary: either an area is wild land or it is not.
The proportion of each agricultural parish designated as wild land was calculated and these results are mapped below. Wild land areas are largely confined to North West Scotland.

```{r SNH_wild_land}
plot.map("wildland_prop", "Wild land\nProportion")
```

### Protected areas

SNH's National Nature Reserve (NNR) data were used as a proxy for hill farming areas, because NNRs are often upland areas, e.g. Beinn Eighe.
Like wild land, they are binary so the proportion of each agricultural parish classified as an NNR was calculated and mapped (below).

```{r SNH_NNR}
plot.map("NNR_prop", "NNR\nProportion")
```

### Carbon and peatland

SNH produce a map of carbon rich soils and peatland, these are polygon data with each polygon taking a categorical variable.
These classifications were split into those which might indicate hill farming and those which do not.
These classifications are listed below:

```{r SNH_peat_areas}
x = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/PEAT_SCOTLAND_parishes.csv") %>% 
   select(a_PRIMARY_LA) %>% 
   distinct()

y=c("Blanket bog/peat. veg.", "Water", "Montane veg.", "Cliffs", "Dubh lochans", "Undiff. heather moor", "Other peat", "Wet heather moor", "Dry heather moor", "Quarries", "Wetlands", "Dune lands", "Bings (area)", "Smooth grass/rushes", "Undiff. Nardus/Molinia", "Snow cover", "Industrial peat", "Ski tows", "Rhododendron")

data.frame(peat_areas=y) %>% 
   kable(col.names="Hill farming classes")

data.frame(not_peat_areas=x$a_PRIMARY_LA[!x$a_PRIMARY_LA %in% y]) %>%
   kable(col.names="Not hill farming classes")
```

Many of these classifications do not obviously relate to the peat or carbon areas, e.g. cloud cover.
Those believed to indicate hill farming constraints were used to calculate the proportion of agricultural parish within them and are plotted below.
North West Scotland appears as the highest proportion using these measures.

```{r SNH_peat}
plot.map("peat_prop", "Peat\nProportion")
```


### JHI LCA Land capability for agriculture

```{r JHI_lca}
plot.map("land_cap_ag_prop", "Agriculture\n6 & 7 class\nProportion")
```

### JHI LCA Land capability for forestry

```{r JHI_lcf}
plot.map("land_cap_forest_prop", "Forestry\nF6 & F7\nProportion")
```


## Group parishes into similar levels

How related are proportions?

```{r proportion_pairs, fig.height=10, fig.width=10}
restrictions %>% 
   select(elev__average, slope__average, grazing_prop, landscape_class_prop, land_cap_ag_prop, land_cap_forest_prop, wildland_prop, rural_prop, peat_prop, NNR_prop) %>% 
   pairs(col=rgb(0, 75/255, 35/255, 0.25),
         upper.panel=panel.cor,
         main="Hilliness indicator comparisons")
```

### Build k-means model

```{r kmean build}
x = restrictions %>% 
   select(elev__average, slope__average, grazing_prop, landscape_class_prop, land_cap_ag_prop, land_cap_forest_prop, wildland_prop, rural_prop, NNR_prop, peat_prop)

# Standardise values
x = scale(x)

# Determine number of clusters
wss = (nrow(x) -1) * sum(apply(x, 2, var))
for (i in 2:10){
   wss[i] = sum(kmeans(x, centers=i)$withinss)
}

plot(1:10, wss, type="b", main="Required number of clusters", xlab="Number of Clusters", ylab="Within groups sum of squares")
```

```{r kmean run}
set.seed(224)

fit2 = kmeans(x, 2) # 2 cluster solution

# append cluster assignment
restrictions = data.frame(restrictions, cluster2=fit2$cluster)
```

```{r kmean_map}
parishes = parishes %>% 
   left_join(select(restrictions, PARCode, cluster2))

parishes %>% 
   ggplot(aes(long, lat, group=group)) +
   geom_polygon(aes(fill=as.character(cluster2))) +
   geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
   coord_equal() +
   scale_fill_brewer(palette="Dark2") +
   labs(fill="Hill or not") +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```


## Rank/score each parish in terms of hilliness

Create a score, which can then be ranked.

```{r score}
score = restrictions %>% 
   select(-PARName,
          -elev__minimum, -elev__median, -elev__maximum,
          -slope__minimum, -slope__median, -slope__maximum,
          -SPA_prop) %>% 
   mutate(elev__average=elev__average/max(elev__average),
          slope__average=slope__average/max(slope__average)) %>% 
   gather(variable, value, -PARCode) %>% 
   group_by(PARCode) %>% 
   summarise(score=sum(value))

parishes = parishes %>% 
   left_join(score)

parishes %>% 
   ggplot(aes(long, lat, fill=score, group=group)) +
   geom_polygon() +
   coord_equal() +
   scale_fill_viridis(option="magma") +
   labs(fill="Hill\nscore") +
   theme_minimal() +
   theme(axis.text=element_blank(),
         axis.title=element_blank(),
         line=element_blank())
```

