---
title: "Census employment data QA"
author: "Mike Spencer"
date: "3 July 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F)
library(rgdal)
library(broom)
library(tidyverse)
library(patchwork)
library(viridis)
library(knitr)
options(scipen = 999)
```


## Introduction

### Purpose

This document is part of a project evaluating the importance of hill farming in Scottish communities. It compares census data at output area level (as downloaded), to aggregations at agricultural parish level.


```{r agricultural_parishes, fig.cap="2016 agricultural parishes"}
parishes = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "simplified_parishes")

Scotland = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "Scotland") %>% 
   tidy()

parishes$id = row.names(parishes)
parishes = tidy(parishes) %>% 
   left_join(parishes@data)
```

```{r data load}
census = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/pop_census.csv") %>% 
   mutate_all(funs(replace(., is.na(.), 0)))

parishes = parishes %>% 
   left_join(census, by=c("PARCode"="PARCode", "PARName"="PARName")) %>% 
   mutate_all(funs(replace(., is.na(.), 0)))
```


```{r functions}
plot.map = function(i, tit){
   ggplot(parishes, aes(long, lat, group=group)) +
      geom_polygon(aes_string(fill=i)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

prop.map = function(i, j, tit){
   parishes %>% 
      rename_(x=i, y=j) %>% 
      mutate(plot_col=x / y) %>% 
   ggplot(aes(long, lat, group=group)) +
      geom_polygon(aes(fill=plot_col)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

panel.cor = function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr = par("usr"); on.exit(par(usr))
    par(usr=c(0, 1, 0, 1))
    r = cor(x, y)
    txt = format(c(r, 0.123456789), digits = digits)[1]
    txt = paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor = 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```

## Absolute values for agricultural parishes

```{r pop totals}
plot.map("pop_all_2011", "2011\npopulation")

plot.map("pop_working_all_2011", "2011 working\npopulation")

plot.map("pop_working_land_2011", "2011 working in\nprimary production\npopulation")
```

Note that these results may be skewed by an issue with island polygons.

## Proportions for agricultural parishes

Proportions should correct nearly all potential issues with island polygons (until I fix the underlying cause), because populations are elevated for all areas. So taking relative values (proportions) 

```{r pop proportions}
prop.map("pop_working_land_2011", "pop_all_2011", "Proportion of\ntotal population\nemployed in\nprimary production")

prop.map("pop_working_land_2011", "pop_working_all_2011", "Proportion of\nworking population\nemployed in\nprimary production")
```

## Checks

### Aggregated to agricultural parishes

* Total people in 2011 = `r round(sum(census$pop_all_2011))`
* Total working population in 2011 = `r round(sum(census$pop_working_all_2011))`
* Ag/for/fish working population in 2011 = `r round(sum(census$pop_working_land_2011))`
* Proportion in land/primary of total population in 2011 = `r scales::percent(sum(census$pop_working_land_2011) / sum(census$pop_all_2011))`
* Proportion in land/primary of working population in 2011 = `r scales::percent(sum(census$pop_working_land_2011) / sum(census$pop_working_all_2011))`

### Using census output areas

These values are the benchmark, as they are unaltered from the downloaded versions.

```{r census OAs}
census_OA_2001 = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/census_OA_2001.csv") %>%
   filter(TAG!="Scotland")
census_OA_2011 = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/census_OA_2011.csv") %>%
   filter(census_id!="Scotland")

census_spatial = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/", "OutputArea2011_MHW.shp"))
   
census_spatial$id = row.names(census_spatial)
census_spatial = tidy(census_spatial) %>% 
   left_join(census_spatial@data) %>% 
   select(long, lat, order, hole, piece, group, OBJECTID, code) %>% 
   left_join(census_OA_2011, by=c("code"="census_id"))
```

* Total people in 2011 = `r sum(census_OA_2011$pop_all_2011, na.rm=T)`
* Total working population in 2011 = `r sum(census_OA_2011$pop_working_all_2011, na.rm=T)`
* Ag/for/fish working population in 2011 = `r sum(census_OA_2011$pop_working_land_2011, na.rm=T)`
* Proportion in land/primary of total population in 2011 = `r scales::percent(sum(census_OA_2011$pop_working_land_2011, na.rm=T) / sum(census_OA_2011$pop_all_2011, na.rm=T))`
* Proportion in land/primary of working population in 2011 = `r scales::percent(sum(census_OA_2011$pop_working_land_2011, na.rm=T) / sum(census_OA_2011$pop_working_all_2011, na.rm=T))`

* Total people in 2001 = `r sum(census_OA_2001$pop_all_2001, na.rm=T)`
* Total working population in 2001 = `r sum(census_OA_2001$pop_working_all_2001, na.rm=T)`
* Ag/for/fish working population in 2001 = `r sum(census_OA_2001$pop_working_land_2001, na.rm=T)`
* Proportion in land/primary of total population in 2001 = `r scales::percent(sum(census_OA_2001$pop_working_land_2001, na.rm=T) / sum(census_OA_2001$pop_all_2001, na.rm=T))`
* Proportion in land/primary of working population in 2001 = `r scales::percent(sum(census_OA_2001$pop_working_land_2001, na.rm=T) / sum(census_OA_2001$pop_working_all_2001, na.rm=T))`


Aggregation is causing inflation in certain areas. Likely due to rounding error, where proportions of a person are increased to the nearest whole number. These errors are very small across the whole of Scotland, but may have impact in isolated parishes.

### Differences between ag parishes and census OA

Using the census data as a benchmark, census data have been subtracted from aggregated data. Hence negative values mean aggregated data are less than the census output area data.

* Total people in 2011 = `r round(sum(census$pop_all_2011) - sum(census_OA_2011$pop_all_2011))`
* Total working population in 2011 = `r round(sum(census$pop_working_all_2011) - sum(census_OA_2011$pop_working_all_2011))`
* Ag/for/fish working population in 2011 = `r round(sum(census$pop_working_land_2011) - sum(census_OA_2011$pop_working_land_2011))`

## Maps of census output areas

```{r census map}
ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Population\nworking on\nthe land (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())

ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011/pop_all_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Proportion of\npopulation\nin primary\nproduction (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())

ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011/pop_working_all_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Proportion of\nworking population\nin primary\nproduction (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
```


## Compared to historic parishes

Used here are employment figures.

```{r parish data}
civ_parish_2011 = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/civil_parish/KS605SC.csv", skip=4, na="-") %>% 
   rename(parish=X1,
          all=`All people aged 16 to 74 in employment`,                             
          A_agri=`A. Agriculture, forestry and fishing`,
          B_mining=`B. Mining and quarrying`,
          C_manufacture=`C. Manufacturing`,
          D_utility=`D. Electricity, gas, steam and air conditioning supply`,
          E_water=`E. Water supply, sewerage, waste management and remediation activities`,
          F_construction=`F. Construction`,
          G_retail=`G. Wholesale and retail trade, repair of motor vehicles and motorcycles`,
          H_transport=`H. Transport and storage`,
          I_accommo=`I. Accommodation and food service activities`,
          J_IT=`J. Information and communication`,
          K_finance=`K. Financial and insurance activities`,
          L_estate=`L. Real estate activities`,
          M_science=`M. Professional, scientific and technical activities`,
          N_admin=`N. Administrative and support service activities`,
          O_public=`O. Public administration and defence, compulsory social security`,
          P_edu=`P. Education`,
          Q_health=`Q. Human health and social work activities`,
          other=`R, S, T, U. Other`) %>% 
   mutate_all(funs(replace(., is.na(.), 0))) %>% 
   mutate(A_agri=100*A_agri/all,
          B_mining=100*B_mining/all,
          C_manufacture=100*C_manufacture/all,
          D_utility=100*D_utility/all,
          E_water=100*E_water/all,
          F_construction=100*F_construction/all,
          G_retail=100*G_retail/all,
          H_transport=100*H_transport/all,
          I_accommo=100*I_accommo/all,
          J_IT=100*J_IT/all,
          K_finance=100*K_finance/all,
          L_estate=100*L_estate/all,
          M_science=100*M_science/all,
          N_admin=100*N_admin/all,
          O_public=100*O_public/all) %>% 
   filter(parish!="Scotland")

civ_parish_2001 = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/civil_parish/KS11a.csv", skip=5, na="-") %>% 
   transmute(parish=X1,
          all=`All people aged 16 to 74 in employment`,                             
          A_agri=`Percentage of people aged 16 to 74 working in: Agriculture, hunting and forestry` + `Percentage of people aged 16 to 74 working in: Fishing`,
          B_mining=`Percentage of people aged 16 to 74 working in: Mining and quarrying`,
          C_manufacture=`Percentage of people aged 16 to 74 working in: Manufacturing`,
          D_utility=`Percentage of people aged 16 to 74 working in: Electricity, gas and water supply`,
          F_construction=`Percentage of people aged 16 to 74 working in: Construction`,
          G_retail=`Percentage of people aged 16 to 74 working in: Wholesale and retail trade, repairs`,
          H_transport=`Percentage of people aged 16 to 74 working in: Transport, storage and communications`,
          I_accommo=`Percentage of people aged 16 to 74 working in: Hotels and restaurants`,
          K_finance=`Percentage of people aged 16 to 74 working as: Financial intermediaries`,
          L_estate=`Percentage of people aged 16 to 74 working in: Real estate, renting and business activities`,
          O_public=`Percentage of people aged 16 to 74 working in: Public administration and defence, social security`,
          P_edu=`Percentage of people aged 16 to 74 working in: Education`,
          Q_health=`Percentage of people aged 16 to 74 working in: Health and social work`,
          other=`Percentage of people aged 16 to 74 working in: Other`) %>% 
   mutate_all(funs(replace(., is.na(.), 0))) %>% 
   filter(parish!="Scotland")

civ_parish = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial"), "CivilParish1930")

civ_parish$id = row.names(civ_parish)
civ_parish = tidy(civ_parish) %>% 
   left_join(civ_parish@data)
```

### 2001 and 2011 census sic match

Not all census sic codes match, this is how they have been joined:

* A_agri=`Percentage of people aged 16 to 74 working in: Agriculture, hunting and forestry` + `Percentage of people aged 16 to 74 working in: Fishing`,
* B_mining=`Percentage of people aged 16 to 74 working in: Mining and quarrying`,
* C_manufacture=`Percentage of people aged 16 to 74 working in: Manufacturing`,
* F_construction=`Percentage of people aged 16 to 74 working in: Construction`,
* G_retail=`Percentage of people aged 16 to 74 working in: Wholesale and retail trade, repairs`,
* I_accommo=`Percentage of people aged 16 to 74 working in: Hotels and restaurants`,
* K_finance=`Percentage of people aged 16 to 74 working as: Financial intermediaries`,
* L_estate=`Percentage of people aged 16 to 74 working in: Real estate, renting and business activities`,
* O_public=`Percentage of people aged 16 to 74 working in: Public administration and defence, social security`,
* P_edu=`Percentage of people aged 16 to 74 working in: Education`,
* Q_health=`Percentage of people aged 16 to 74 working in: Health and social work`,
* other=`Percentage of people aged 16 to 74 working in: Other`)

### Civil parish change

How has employment changed in civil parishes between the 2001 and 2011 census?

```{r civ parish change}
civ_parish_2001 %>% 
   select(parish, all) %>% 
   inner_join(select(civ_parish_2011, parish, all), by=c("parish"="parish")) %>% 
   mutate(diff=all.x/all.y) %>% 
   ggplot(aes(diff)) +
   geom_histogram() +
   scale_x_continuous(labels=scales::percent) +
   labs(title="Employment in civil parishes (aged 16-74)",
        subtitle="Histogram of percentage change from 2001 to 2011",
        x="",
        y="Number of parishes")
```

```{r civil functions}
plot.civ=function(colu, tit){
   x = civ_parish_2001 %>% 
      inner_join(civ_parish_2011, by=c("parish"="parish")) %>% 
      rename_(col2001=paste0(colu, ".x"),
              col2011=paste0(colu, ".y")) %>% 
   transmute(diff=(col2011-col2001)/100)
   
   civ_parish_2001 %>% 
      inner_join(civ_parish_2011, by=c("parish"="parish")) %>% 
      rename_(col2001=paste0(colu, ".x"),
              col2011=paste0(colu, ".y")) %>% 
      transmute(diff=(col2011-col2001)/100) %>% 
      ggplot(aes(diff)) +
      geom_histogram() +
      scale_x_continuous(labels=scales::percent) +
      labs(title=tit,
           subtitle=paste("Percentage change from 2001 to 2011 |",
                          sum(x>0),
                           "of",
                          nrow(x),
                           "parishes increase"),
           x="",
           y="Number of parishes")   
}

map.civ = function(colu, tit){
   x = civ_parish_2001 %>% 
   inner_join(civ_parish_2011, by=c("parish"="parish")) %>% 
   rename_(col2001=paste0(colu, ".x"),
           col2011=paste0(colu, ".y")) %>% 
   mutate(diff=(col2011-col2001)/100) %>% 
   select(parish, diff)

civ_parish %>% 
   left_join(x, by=c("name"="parish")) %>% 
   ggplot(aes(long, lat, group=group)) +
      geom_polygon(aes(fill=diff)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(type="div",
                           direction=1,
                           labels=scales::percent,
                           limits=c(-0.3, 0.3)) +
      labs(title=tit,
           fill="Percentage\nchange") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}
```

```{r civ parish change sector}
plot.civ("A_agri", "Employed in agriculture, forestry or fishing") +
   map.civ("A_agri", "")

plot.civ("B_mining", "Employed in mining and quarrying") +
   map.civ("B_mining", "")

plot.civ("C_manufacture", "Employed in manufacturing") +
   map.civ("C_manufacture", "")

plot.civ("F_construction", "Employed in construction") +
   map.civ("F_construction", "")

plot.civ("G_retail", "Employed in wholesale and retail trade, repairs") +
   map.civ("G_retail", "")

plot.civ("I_accommo", "Employed in hotels and restaurants") +
   map.civ("I_accommo", "")

plot.civ("K_finance", "Employed in finance") +
   map.civ("K_finance", "")

plot.civ("L_estate", "Employed in real estate") +
   map.civ("L_estate", "")

plot.civ("O_public", "Employed in public sector") +
   map.civ("O_public", "")
```

### Comparing agricultural and civic parishes

```{r ag2civ parishes all employment}
x = census %>% 
   inner_join(civ_parish_2001, by=c("PARName"="parish")) %>% 
   inner_join(civ_parish_2011, by=c("PARName"="parish"))

x %>% 
   mutate(diff2001=(pop_working_all_2001 - all.x)) %>% 
   ggplot(aes(pop_working_all_2001, all.x)) +
   geom_point() +
   geom_abline(slope=1) +
   geom_text(aes(label=ifelse(!between(diff2001, -1000, 1000), 
                              as.character(PARName), "")),
             hjust=0, vjust=0) +
   labs(title="Those aged 16-74 in employment by parish",
        subtitle="2001",
        x="Aggregated to agricultural parishes",
        y="Downloaded for civic parishes") +
   ggplot(x, aes(pop_working_all_2011, all.y)) +
   geom_point() +
   geom_abline(slope=1) +
   labs(title="",
        subtitle="2011",
        x="Aggregated to agricultural parishes",
        y="")
```

There are two agricultural parishes called Cathcart, when their populations are added together they sum to the population of the civic parish with the same name.

Total working population in 2011 for aggregated agricultural parishes is `r sum(census$pop_working_all_2011)` people and for civic parishes `r sum(civ_parish_2011$all)` people.
In 2001 the total working population for aggregated agricultural parishes is `r sum(census$pop_working_all_2001)` people and for civic parishes `r sum(civ_parish_2001$all)` people.

```{r ag2civ parishes land employment}
x = census %>% 
   inner_join(civ_parish_2001, by=c("PARName"="parish")) %>% 
   inner_join(civ_parish_2011, by=c("PARName"="parish"))

y = x %>% 
   mutate(prop2001=100 * pop_working_land_2001/pop_working_all_2001,
          prop2011=100 * pop_working_land_2011/pop_working_all_2011,
          diff2001=(prop2001 - A_agri.x))

ggplot(y, aes(prop2001, A_agri.x)) +
   geom_point() +
   geom_abline(slope=1) +
   geom_text(aes(label=ifelse(!between(diff2001, -1000, 1000), 
                              as.character(PARName), "")),
             hjust=0, vjust=0) +
   labs(title="Percentage of those aged 16-74 in agriculture, forestry or fishing employment by parish",
        subtitle="2001",
        x="Aggregated to agricultural parishes",
        y="Downloaded for civic parishes") +
   ggplot(y, aes(prop2011, A_agri.y)) +
   geom_point() +
   geom_abline(slope=1) +
   labs(title="",
        subtitle="2011",
        x="Aggregated to agricultural parishes",
        y="")
```

SIC code A from 2011 was used as a base, but was different in 2001. I have aggregated two 2001 employment categories to create one broadly comparable to the SIC A of 2011.
The working in agriculture/forestry/fishing population for aggregated parishes in 2011 is `r sum(census$pop_working_land_2011)` people and for civic parishes `r sum(civ_parish_2011$A_agri/100 * civ_parish_2011$all, na.rm=T)` people.
compare them to the census output area value, which is `r sum(census_OA_2011$pop_working_land_2011, na.rm=T)` people.
The same values for 2001 are: aggregated agricultural parishes, `r sum(census$pop_working_land_2001)`; civic parishes `r sum(civ_parish_2001$A_agri/100 * civ_parish_2001$all, na.rm=T)` people; `r sum(census_OA_2001$pop_working_land_2001)` people.
The largest differences are between the civic parishes and the output areas/aggregated parishes for 2001.

