---
title: "Hill farming and employment"
author: "Mike Spencer"
date: "3 July 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F)
library(rgdal)
library(broom)
library(tidyverse)
library(viridis)
library(knitr)
```


## Introduction

### Purpose

This document is part of a project evaluating the importance of hill farming in Scottish communities.

### Output level

Results will be aggregated to agricultural parishes, as this administrative level is widely understood and is appropriate for anonymising data.
The 2016 agricultural parish data were downloaded from: <https://data.gov.uk/dataset/939fdd5e-7322-4ab7-9dc9-bbfc538c4477/agricultural-parishes>.

```{r agricultural_parishes, fig.cap="2016 agricultural parishes"}
parishes = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "simplified_parishes")

Scotland = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/ag_parishes_2016.gpkg"), "Scotland") %>% 
   tidy()

parishes$id = row.names(parishes)
parishes = tidy(parishes) %>% 
   left_join(parishes@data)
```

```{r data load}
restrictions = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/spatial-processed/parish_restrictions.csv") %>% 
   mutate_all(funs(replace(., is.na(.), 0)))

census = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/pop_census.csv") %>% 
   mutate_all(funs(replace(., is.na(.), 0)))

parishes = parishes %>% 
   left_join(census, by=c("PARCode"="PARCode", "PARName"="PARName")) %>% 
   mutate_all(funs(replace(., is.na(.), 0)))
```


```{r functions}
plot.map = function(i, tit){
   ggplot(parishes, aes(long, lat, group=group)) +
      geom_polygon(aes_string(fill=i)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

prop.map = function(i, j, tit){
   parishes %>% 
      rename_(x=i, y=j) %>% 
      mutate(plot_col=x / y) %>% 
   ggplot(aes(long, lat, group=group)) +
      geom_polygon(aes(fill=plot_col)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill=tit) +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
}

panel.cor = function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr = par("usr"); on.exit(par(usr))
    par(usr=c(0, 1, 0, 1))
    r = cor(x, y)
    txt = format(c(r, 0.123456789), digits = digits)[1]
    txt = paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor = 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```

## Absolute values

```{r pop totals}
plot.map("pop_all_2011", "2011\npopulation")

plot.map("pop_working_all_2011", "2011 working\npopulation")

plot.map("pop_working_land_2011", "2011 working in\nprimary production\npopulation")
```

Note that these results are skewed by an issue with island polygons.

## Proportions

Proportions should correct nearly all issues with island polygons (until I fix the underlying cause).

```{r pop proportions}
prop.map("pop_working_land_2011", "pop_all_2011", "Proportion of\ntotal population\nemployed in\nprimary production")

prop.map("pop_working_land_2011", "pop_working_all_2011", "Proportion of\nworking population\nemployed in\nprimary production")
```

## Checking!

### Aggregated to agricultural parishes

* Total people in 2011 = `r sum(census$pop_all_2011)`
* Proportion in land/primary of total population in 2011 = `r scales::percent(sum(census$pop_working_land_2011) / sum(census$pop_all_2011))`
* Proportion in land/primary of working population in 2011 = `r scales::percent(sum(census$pop_working_land_2011) / sum(census$pop_working_all_2011))`

### Using raw census output areas

```{r census OAs}
census_OA = read_csv("~/Cloud/Michael/SRUC/hill_farms/data/census/census2011.csv") %>%
   filter(census_id!="Scotland")

census_spatial = readOGR(paste0(normalizePath("~"), "/Cloud/Michael/SRUC/hill_farms/data/spatial/", "OutputArea2011_MHW.shp"))
   
census_spatial$id = row.names(census_spatial)
census_spatial = tidy(census_spatial) %>% 
   left_join(census_spatial@data) %>% 
   select(long, lat, order, hole, piece, group, OBJECTID, code) %>% 
   left_join(census_OA, by=c("code"="census_id"))
```

* Total people in 2011 = `r sum(census_OA$pop_all_2011)`
* Proportion in land/primary of total population in 2011 = `r scales::percent(sum(census_OA$pop_working_land_2011) / sum(census_OA$pop_all_2011))`
* Proportion in land/primary of working population in 2011 = `r scales::percent(sum(census_OA$pop_working_land_2011) / sum(census_OA$pop_working_all_2011))`

Aggregation is causing inflation in certain areas. Likely due to rounding error, where proportions of a person are increased to the next whole number.

```{r census map}
ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Population\nworking on\nthe land (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())

ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011/pop_all_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Proportion of\npopulation\nin primary\nproduction (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())

ggplot(census_spatial, aes(long, lat, group=group)) +
      geom_polygon(aes(fill=pop_working_land_2011/pop_working_all_2011)) +
      geom_polygon(data=Scotland, aes(long, lat, group=group),
                   colour="grey30", fill=NA, size=0.1) +
      coord_equal() +
      scale_fill_distiller(palette="Greens", direction=1) +
      labs(fill="Proportion of\nworking population\nin primary\nproduction (2011)") +
      theme_minimal() +
      theme(axis.text=element_blank(),
            axis.title=element_blank(),
            line=element_blank())
```

